Stop Lying to Me: New Visual Tools to Choose the Most Honest Nonlinear Dimension Reduction

### 2025-06-26 appendix.qmd
Updated appendix Figure 5 title

```{r}
#| label: fig-pbmc-mse-umap
#| fig-pos: H
#| out-width: 100%
#| fig-cap: "Assessing which of the two layouts with UMAP and tSNE with different hyper-parameter setting (n_neighbors: $30$, min_dist: $0.3$ (red); n_neighbors: $30$ (blue)) on the PBMC data is the better representation using RMSE for varying binwidth ($a_1$). Colour used for the lines and points in the top plot and in the scatterplots represents UMAP and tSNE layouts (a, b). Of the two, layout b is optimal across all binwidths making it the best choice."

@@

```{r}
#| echo: false
#| fig-cap: "Compare the published \\gD{} layout (@fig-pbmc-mse-umap a) and the \\gD{} layout selected (@fig-pbmc-mse-umap b) by RMSE plot (@fig-pbmc-mse-umap) from tSNE and UMAP with default hyper-parameters. The PBMC data ($n =  5858$) has three clusters in $9\\text{-}D$, where three clusters are close. Two \\gD{} projection from a tour on $9\\text{-}D$ of the model fit with @fig-pbmc-mse-umap a ($a_1 = 0.04$, $b = 900/179 (30, 30)$) shows three-well separated clusters with big separations. On the other hand, the model fit with @fig-pbmc-mse-umap b ($a_1 = 0.04$, $b = 1110/317 (30, 37)$) shows three-well separated clusters with small separations. Therefore, @fig-pbmc-mse-umap b is more reasonable than @fig-pbmc-mse-umap a. Videos of the langevitour animations are available at <https://youtu.be/0cKX_HG_n0k> and <https://youtu.be/KhJvsRtaX04> respectively."
#| label: fig-model-pbmc-author-proj
#| fig-pos: H
#| fig-width: 10
#| fig-height: 15
#| out-width: 90%

>>>>

```{r}
#| label: fig-pbmc-mse-umap
#| fig-pos: H
#| out-width: 100%
#| fig-cap: "Assessing which of the 2 NLDR layouts on the PBMC3k data is the better representation using RMSE for varying binwidth ($a_1$). Color used for the lines and points in the left plot and in the scatterplots represents NLDR layouts (a, b). Of the two, layout b is optimal across all binwidths making it the best choice."

@@

```{r}
#| echo: false
#| fig-cap: "Compare the published \\gD{} layout (@fig-pbmc-mse-umap a) made with UMAP and the \\gD{} layout selected (@fig-pbmc-mse-umap b) made with tSNE by RMSE plot (@fig-pbmc-mse-umap). The two plots on the right show  projections from a tour, with the models overlaid. The published layout a suggested three separated clusters with two of them are close, but this is not present in the data. While there may be three clusters they are not well-separated. The difference in model fit also indicates this: the published layout a does not capture the nonlinear structure of the clusters like the model generated from layout b. This supports the choice that layout b is the better representation of the data, because it shows close clusters. Videos of the langevitour animations are available at <https://youtu.be/ffiB4MGWyn8> and <https://youtu.be/e7XNL18co1c> respectively."
#| label: fig-model-pbmc-author-proj
#| fig-pos: H
#| fig-width: 10
#| fig-height: 15
#| out-width: 90%

### 2025-06-26 appendix.tex
Figure 1 link video
Updated appendix Figure 5 title

\caption{Two projections of the \pD{} true model overlaying the data are
shown in a, b.}

@@

\caption{\label{fig-pbmc-mse-umap}Assessing which of the two layouts
with UMAP and tSNE with different hyper-parameter setting (n\_neighbors:
\(30\), min\_dist: \(0.3\) (red); n\_neighbors: \(30\) (blue)) on the
PBMC data is the better representation using RMSE for varying binwidth
(\(a_1\)). Colour used for the lines and points in the top plot and in
the scatterplots represents UMAP and tSNE layouts (a, b). Of the two,
layout b is optimal across all binwidths making it the best choice.}

@@

\caption{\label{fig-model-pbmc-author-proj}Compare the published \gD{}
layout (Figure~\ref{fig-pbmc-mse-umap} a) and the \gD{} layout selected
(Figure~\ref{fig-pbmc-mse-umap} b) by RMSE plot
(Figure~\ref{fig-pbmc-mse-umap}) from tSNE and UMAP with default
hyper-parameters. The PBMC data (\(n =  5858\)) has three clusters in
\(9\text{-}D\), where three clusters are close. Two \gD{} projection
from a tour on \(9\text{-}D\) of the model fit with
Figure~\ref{fig-pbmc-mse-umap} a (\(a_1 = 0.04\),
\(b = 900/179 (30, 30)\)) shows three-well separated clusters with big
separations. On the other hand, the model fit with
Figure~\ref{fig-pbmc-mse-umap} b (\(a_1 = 0.04\),
\(b = 1110/317 (30, 37)\)) shows three-well separated clusters with
small separations. Therefore, Figure~\ref{fig-pbmc-mse-umap} b is more
reasonable than Figure~\ref{fig-pbmc-mse-umap} a. Videos of the
langevitour animations are available at
\url{https://youtu.be/0cKX_HG_n0k} and
\url{https://youtu.be/KhJvsRtaX04} respectively.}

>>>>

\caption{Two projections of the \pD{} true model overlaying the data are
shown in a, b. Video of the langevitour animations is available at
\url{https://youtu.be/35TrnYJsUUI}.}

@@

\caption{\label{fig-pbmc-mse-umap}Assessing which of the 2 NLDR layouts
on the PBMC3k data is the better representation using RMSE for varying
binwidth (\(a_1\)). Color used for the lines and points in the left plot
and in the scatterplots represents NLDR layouts (a, b). Of the two,
layout b is optimal across all binwidths making it the best choice.}

@@

\caption{\label{fig-model-pbmc-author-proj}Compare the published \gD{}
layout (Figure~\ref{fig-pbmc-mse-umap} a) made with UMAP and the \gD{}
layout selected (Figure~\ref{fig-pbmc-mse-umap} b) made with tSNE by
RMSE plot (Figure~\ref{fig-pbmc-mse-umap}). The two plots on the right
show projections from a tour, with the models overlaid. The published
layout a suggested three separated clusters with two of them are close,
but this is not present in the data. While there may be three clusters
they are not well-separated. The difference in model fit also indicates
this: the published layout a does not capture the nonlinear structure of
the clusters like the model generated from layout b. This supports the
choice that layout b is the better representation of the data, because
it shows close clusters. Videos of the langevitour animations are
available at \url{https://youtu.be/ffiB4MGWyn8} and
\url{https://youtu.be/e7XNL18co1c} respectively.}

### 2025-06-26 paper.qmd
Updated Figure 15
Updated appendix Figure 5 title

```{r}
#| echo: false
#| fig-cap: "Compare the published \\gD{} layout (a) made with UMAP and the \\gD{} layout selected by RMSE plot (e) made by tSNE. The two plots on the right show  projections from a tour, with the models overlaid. The published layout suggested three very separated clusters, but this is not present in the data. While there may be three clusters they are not well-separated. The difference in model fit also indicates this: the published layout a does not spread out fully into the point cloud like the model generated from layout e. This supports the choice that layout e is the better representation of the data, because it does not exaggerate separation between clusters."
#| label: fig-model-pbmc-author-proj
#| fig-pos: H
#| fig-width: 10
#| fig-height: 15
#| out-width: 90%

@@

imge_error_sample_inside <- plot_digit_img(
  img_error_inside, color_on = "#ff7f00", title_text = "f")

>>>>

```{r}
#| echo: false
#| fig-cap: "Compare the published \\gD{} layout (a) made with UMAP and the \\gD{} layout selected by RMSE plot (e) made by tSNE. The two plots on the right show  projections from a tour, with the models overlaid. The published layout a suggested three very separated clusters, but this is not present in the data. While there may be three clusters they are not well-separated. The difference in model fit also indicates this: the published layout a does not spread out fully into the point cloud like the model generated from layout e. This supports the choice that layout e is the better representation of the data, because it does not exaggerate separation between clusters."
#| label: fig-model-pbmc-author-proj
#| fig-pos: H
#| fig-width: 10
#| fig-height: 15
#| out-width: 90%

@@

imge_error_sample_inside <- plot_digit_img(
  img_error_inside, color_on = "#d95f0e", title_text = "f")
